Rheology of Cubic Blue Phases

### 2013-07-04 trunk/papers/bprheo/bprheo.tex
Update bprheo paper -- included Mike's comments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2041 40876441-3496-0410-a613-a2fca35b2ae2

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
}
Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- 
\rev{
very slow flow causes a transition of the initially 
ordered structure with oscillatory response into an amorphous network 
with an apparent yield stress. 
Larger shear rates lead to another amorphous state with different structure of 
the defect network.
}
For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, 
and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. \rev{
Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
}
Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
\rev{
However, this result was derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.
}  
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma}\, l$. 
\rev{
The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
}

The system of coupled partial differential equations~\ref{op-eom}
and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
which uses a combination of lattice Boltzmann and finite difference schemes.
(This is in contrast with some earlier methods using solely
LB~\cite{Denniston:2001, Denniston:2004}.)
The Navier-Stokes
equation is solved via the lattice Boltzmann approach, using a standard
three-dimensional model with 19 discrete velocities (D3Q19).
A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
used and the time step is $\Delta t = 1$ in lattice units.
Coupling to the thermodynamic sector is via a
local body force computed as the divergence of the thermodynamic
stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
of the time evolution of $\mathbf{Q}$ via a standard finite difference
method using the same grid and the same time step as the LB. The system
is periodic in all three coordinate directions. Sliding
boundary conditions, or Lees-Edwards planes ~\cite{Wagner:2002}, are explicitly implemented
in both hydrodynamic and thermodynamic sectors to impose shear on the
system. 
\rev{
The same hybrid method with Lees-Edwards planes
has been used successfully to study binary mixtures, particle suspensions~\cite{MacMeccan:2009, Aidun:2010}, 
emulsions~\cite{Frijters:2012} and smectic liquid crystals~\cite{Henrich:2012a}.
A more detailed discussion of our use of Lees-Edwards boundary conditions can be found in the Appendix.
}

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than three orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $3.75\times\e{-3}$ LBU.
\rev{
For clarity we classify various flow regimes, namely three in the case of Blue Phase II and five in the 
case of Blue Phase I. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) which 
may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
GJ and FAN occur in both blue phases as steady states regardless of the initial state.
}

@@

\begin{table}[htpb]
\begin{tabular}{|c|c|c|}
\hline
Name & $\gd$ & $Er$ \\
\hline
BPI-1 (AN) & $ \lesssim 1.95\e{-5}$ & $\lesssim 0.17$ \\
BPI-2 (PRC) & $3.91\e{-5} \dots 2.34\e{-4}$ & $0.33 \dots 2$ \\
BPI-3 (AN) & $3.13\e{-4}\dots 4.69\e{-4}$ & $2.67 \dots 4$ \\
BPI-4 (GJ, FGJ) & $5.47\e{-4} \dots 1.88\e{-3}$ & $4.67 \dots 16$ \\
BPI-5 (FAN) & $\gtrsim 2.5\e{-3}$ & $ \gtrsim 21.33$ \\
\hline
BPII-1 (PRC) & $\lesssim 3.91\e{-4}$ & $ \lesssim 3.33$\\
BPII-2 (GJ) & $4.69\dots 1.25\e{-3}$ & $4 \dots 10.67$ \\
BPII-3 (FAN) &$\gtrsim 1.88\e{-3}$ & $ \gtrsim 16$ \\ 
\hline
\end{tabular}
\label{table2}
\end{table}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
\caption{
\rev{
Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
}
}
\label{bp2-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{
\rev{
Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
of BPII versus strain $\gamma = \gd\cdot t$  
normalised to the background viscosity $\eta$.
The pictures show the regimes BPII-1 (periodically recurring conformations, top), BPII-2 (Grandjean texture, bottom) and BPII-3 (flow-aligned nematic, bottom). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. total shear-thinning.
}
} 
\label{bp2-appvisc}
\end{figure}

@@

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
\rev{
Time averages were taken as average over one stress cycle in the steady state.
Where no oscillations occurred we used the final value at the end of the run.
For the lowest up to intermediate shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure with the helical axis oriented along the
flow gradient direction ($y$). The coefficients of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, for the range of shear rates which we have explored here, 
the degree of shear-thinning is remarkably small in BPII.
}

\rev{
Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional stress contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic break-up and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence of a permeation mode at very low shear rates may be due to our
choice of boundary conditions. We refer the reader to the Appendix for a detailed discussion.
}

\rev{
We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director predominantly oriented in flow-vorticity plane and having only
a small angle with the flow direction.
The picture at the bottom of Fig.~\ref{bp2-appvisc} shows data for this regime, to which we refer as BPII-2. The 
curves for the two largest flow rates are for BPII-3, the flow-aligned nematic state. 
We will address these regimes in section \ref{gj-fan}, as they are common to both cubic blue phases BPI and BPII.
}

\rev{
The travelling helical wave which was identified in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be a metastable flow state as it features larger free energies, higher stresses and dissipation due 
to the tumbling motion of the director field, which are completely absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
}

@@

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. 
\rev{
Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The secondary flow in vorticity direction takes clearly place in positive {\em and} 
negative z-direction, whereas the movement of the network as a whole is only in one direction.
}

@@

\rev{
Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve} and Fig. \ref{bp1-appvisc} depicts 
the apparent viscosity ratio $\eta_{app}/\eta$ over total strain.
}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{
\rev{
Flow curve $\bar{\Pi}_{xy}(\gd)$: 
Depending on the configuration in steady shear flow five different regimes 
can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
}
\label{bp1-flowcurve}
\end{figure}

For intermediate (but not too low) shear rates
($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more 
find regular oscillations in the network morphology and in the stress
response versus time. We label this regime BPI-2. 
\rev{
The centre picture of Fig.~\ref{bp1-appvisc} shows the corresponding data for the
apparent viscosity ratio versus strain.
}

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
\rev{
For these low shear rates 
($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
the flow curve in Fig. \ref{bp1-flowcurve}
develops a horizontal yield stress branch. 
We refer to this flow regime as BPI-1.
}

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value, 
\rev{
which is in the region of Ericksen numbers $Er \simeq O(1)$.
This is a regime 
($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
we labelled BPI-3. It is characterised by the balance between viscous and elastic forces.
This has an influence on the stability of the disclination network as it 
seems to become increasingly sensitive to small perturbations in the flow field
which couple back to the order parameter dynamics and vice versa. This phenomenon forms 
a possible route towards an instance of ``rheochaos'' ~\cite{rheochaos,Cates:2002}.
}

\rev{
Finally, for shear rates such that
$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
and before the break-up into a flow-aligned nematic occurs 
(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
we observe the same cholesteric Grandjean configuration with the 
helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{gj-fan}). This regime is referred to as BPI-4.
}

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{
\rev{
Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised to the background viscosity $\eta$: 
The top picture shows regime BPI-1 (amorphous network) where
the BP network transforms into an amorphous network featuring a yield stress.
The centre picture shows regimes BPI-2 (periodically recurring conformations) and BPI-3 (amorphous network), 
whereas the picture at the bottom
depicts data of the (partly frustrated) Grandjean texture in regime BPI-4 
and the flow-aligned nematic state, labelled BPI-5 in this context. 
Note that the $\eta_{app}/\eta$ is negative for the lowest 
flow rates as the (negative) contribution of the elastic forces
of the network dominates the total stress.
}
}
\label{bp1-appvisc}
\end{figure}

@@

\begin{figure*}[htpb]
\includegraphics[width=0.32\textwidth]{disc-364k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-366k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-368k_run914r.png}\\
\includegraphics[width=0.32\textwidth]{disc-370k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-372k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-374k_run914r.png}
\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
The sequence shows a typical cycle of shear-induced transformations in the 
steady state at $\gd=1.56\e{-4}$ and time steps 
$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
During every cycle the network also is displaced 
\rev{along the vorticity direction just as BPII
in regime BPII-1, but at a different rate.
}
}
\label{bp1-2-disc}
\end{figure*}

>>>>

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction.
Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- 
very slow flow causes a transition of the initially 
ordered structure into an amorphous network 
with an apparent yield stress. 
Larger shear rates lead to another amorphous state with different structure of 
the defect network.
For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, 
and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. 
Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
Rey~\cite{Rey:1996a, Rey:1996b} theoretically studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
However, this result was derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma}\, l$. 
The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.

The system of coupled partial differential equations~\ref{op-eom}
and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
which uses a combination of lattice Boltzmann and finite difference schemes.
(This is in contrast with some earlier methods using solely
LB~\cite{Denniston:2001, Denniston:2004}.)
The Navier-Stokes
equation is solved via the lattice Boltzmann approach, using a standard
three-dimensional model with 19 discrete velocities (D3Q19).
A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
used and the time step is $\Delta t = 1$ in lattice units.
Coupling to the thermodynamic sector is via a
local body force computed as the divergence of the thermodynamic
stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
of the time evolution of $\mathbf{Q}$ via a standard finite difference
method using the same grid and the same time step as the LB. The system
is periodic in all three coordinate directions. Sliding
boundary conditions, or Lees-Edwards planes ~\cite{Wagner:2002}, are explicitly implemented
in both hydrodynamic and thermodynamic sectors to impose shear on the
system. 
The same method with Lees-Edwards planes
has been used successfully to study binary mixtures, particle suspensions~\cite{MacMeccan:2009, Aidun:2010}, 
emulsions~\cite{Frijters:2012} and smectic liquid crystals~\cite{Henrich:2012a}.
A more detailed discussion of our use of Lees-Edwards boundary conditions can be found in the Appendix.

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than three orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $3.75\times\e{-3}$ LBU.
For clarity we classify various flow regimes, namely three in the case of Blue Phase II and five in the 
case of Blue Phase I. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) which 
may be partly frustrated (FGJ) and a flow-aligned nematic state (FN). 
GJ and FN occur in both blue phases as steady states regardless of the initial state.

@@

\begin{table}[htpb]
\begin{tabular}{|c|c|c|}
\hline
Name & $\gd$ & $Er$ \\
\hline
BPI-1 (AN) & $ \lesssim 1.95\e{-5}$ & $\lesssim 0.17$ \\
BPI-2 (PRC) & $3.91\e{-5} \dots 2.34\e{-4}$ & $0.33 \dots 2$ \\
BPI-3 (AN) & $3.13\e{-4}\dots 4.69\e{-4}$ & $2.67 \dots 4$ \\
BPI-4 (GJ, FGJ) & $5.47\e{-4} \dots 1.88\e{-3}$ & $4.67 \dots 16$ \\
BPI-5 (FN) & $\gtrsim 2.5\e{-3}$ & $ \gtrsim 21.33$ \\
\hline
BPII-1 (PRC) & $\lesssim 3.91\e{-4}$ & $ \lesssim 3.33$\\
BPII-2 (GJ) & $4.69\dots 1.25\e{-3}$ & $4 \dots 10.67$ \\
BPII-3 (FN) &$\gtrsim 1.88\e{-3}$ & $ \gtrsim 16$ \\ 
\hline
\end{tabular}
\label{table2}
\end{table}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
\caption{
Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
}
\label{bp2-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{
Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
of BPII versus strain $\gamma = \gd\cdot t$  
normalised to the background viscosity $\eta$.
The pictures show the regimes BPII-1 (periodically recurring conformations, top), BPII-2 (Grandjean texture, bottom) and BPII-3 (flow-aligned nematic, bottom). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. totally shear-thinned state.
} 
\label{bp2-appvisc}
\end{figure}

@@

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
Time averages were taken as average over one stress cycle in the steady state.
Where no oscillations occurred we used the final value at the end of the run.
For the lowest up to intermediate shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure with the helical axis oriented along the
flow gradient direction ($y$). The coefficients of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, for the range of shear rates which we have explored here, 
the degree of shear-thinning is remarkably small in BPII.

Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional stress contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic break-up and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence of a permeation mode at very low shear rates may be due to our
choice of boundary conditions. We refer the reader to the Appendix for a detailed discussion.

We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director predominantly oriented in flow-vorticity plane and having only
a small angle with the flow direction.
The picture at the bottom of Fig.~\ref{bp2-appvisc} shows data for this regime, to which we refer as BPII-2. The 
curves for the two largest flow rates are for BPII-3, the flow-aligned nematic state. 
We will address these regimes in section \ref{gj-fan}, as they are common to both cubic blue phases BPI and BPII.

The travelling helical wave which was identified in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be a metastable flow state as it features larger free energies, higher stresses and dissipation due 
to the tumbling motion of the director field, which are completely absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.

@@

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. 
Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local flow velocity. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The secondary flow in vorticity direction takes clearly place in positive {\em and} 
negative z-direction, whereas the movement of the network as a whole is only in one direction.

@@

Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve} and Fig. \ref{bp1-appvisc} depicts 
the apparent viscosity ratio $\eta_{app}/\eta$ as a function of total strain.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{
Flow curve $\bar{\Pi}_{xy}(\gd)$: 
Depending on the configuration in steady shear flow five different regimes 
can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares); 
steady flow with periodically recurring patterns (BPI-2, open squares); 
break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles);
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture, as well as frustrated metastable Grandjean-like states that can transform into 
a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles);
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
\label{bp1-flowcurve}
\end{figure}

For intermediate (but not too low) shear rates
($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more 
find regular oscillations in the network morphology and in the stress
response versus time. We label this regime BPI-2. 
The centre picture of Fig.~\ref{bp1-appvisc} shows the corresponding data for the
apparent viscosity ratio versus strain.

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
For these low shear rates 
($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
the flow curve in Fig. \ref{bp1-flowcurve}
develops a horizontal yield stress branch. 
We refer to this flow regime as BPI-1.

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value, 
which is in the region of Ericksen numbers $Er \simeq O(1)$.
This is a regime 
($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
we labelled BPI-3. It is characterised by the balance between viscous and elastic forces.
This has an influence on the stability of the disclination network as it 
seems to become increasingly sensitive to small perturbations in the flow field
which couple back to the order parameter dynamics and vice versa. This phenomenon forms 
a possible route towards an instance of ``rheochaos'' ~\cite{rheochaos,Cates:2002}.

Finally, for shear rates such that
$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
and before the break-up into a flow-aligned nematic occurs 
(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
we observe the same cholesteric Grandjean configuration with the 
helical axis along the gradient direction as for BPII-2. 
For clarity we refer to this regime as BPI-4. It will be discussed 
in a separate section \ref{gj-fan}.

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{
Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised to the background viscosity $\eta$: 
The top picture shows regime BPI-1 (amorphous network) where
the BP network transforms into an amorphous network featuring a yield stress.
The centre picture shows regimes BPI-2 (periodically recurring conformations) and BPI-3 (amorphous network), 
whereas the picture at the bottom
depicts data of the (partly frustrated) Grandjean texture in regime BPI-4 
and the flow-aligned nematic state, labelled BPI-5 in this context. 
Note that the $\eta_{app}/\eta$ is sometimes negative for the lowest 
flow rates as the (negative) contribution of the elastic forces
of the network dominates the total stress.
}
\label{bp1-appvisc}
\end{figure}

@@

\begin{figure*}[htpb]
\includegraphics[width=0.32\textwidth]{disc-364k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-366k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-368k_run914r.png}\\
\includegraphics[width=0.32\textwidth]{disc-370k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-372k_run914r.png}
\includegraphics[width=0.32\textwidth]{disc-374k_run914r.png}
\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
The sequence shows a typical cycle of shear-induced transformations in the 
steady state at $\gd=1.56\e{-4}$ and time steps 
$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
During every cycle the network also is displaced 
along the vorticity direction just as BPII
in regime BPII-1, but at a different rate.
}
\label{bp1-2-disc}
\end{figure*}
